---
output: 
  pdf_document:
    keep_tex: yes
    fig_caption: yes
    number_sections: yes
geometry: margin=2.54cm
title: "Air Quality in Ukraine post Ukraine-Russia Dispute"
subtitle: "Web address for GitHub repository"
author: "Curly Girlies: Shirley Fontanié, Rachel Gordon, and Julia Weinberg"
fontsize: 12pt
mainfont: Times New Roman

---

\newpage
\tableofcontents 
\newpage
\listoftables 
\newpage
\listoffigures 
\newpage

```{r setup, include=FALSE}
# Set your working directory 

getwd() 

# Load your packages  

library(tidyverse)
library(lubridate)
library(ggplot2)

# Set your ggplot theme

ourtheme <- theme_classic(base_size = 14)+ 
  theme(axis.text = element_text(color="black"), 
        legend.position = "top")
theme_set(ourtheme)
  
# Load your datasets

UkraineData <- read.csv("./Data/RawData/Ukraine_AQIData.csv", stringsAsFactors = TRUE)

```

# Rationale and Research Questions

> Research question: How does air quality in various Ukrainian cities differ before and after the Ukraine-Russia dispute?  
*******
> Rationale: On February 24, 2022, Russia attacked Ukraine. The first city attacked was
Lviv and Dnipro. 
> Kyiv was hit February 24th.  

\newpage

# Dataset Information 

> Describe sources of data here (input Julia paragraph)

```{r Data wrangling}  
# Initial wrangling 

Ukraine_Processed <- UkraineData %>% 
  drop_na(pm10) %>% 
  drop_na(pm25) 

# Setting Date 

Ukraine_Processed$date <- as.Date(Ukraine_Processed$date, "%m/%d/%y")

# Creating subsets by city  

Dnipro <- Ukraine_Processed %>% 
  filter(City %in% "Dnipro")

Lviv <- Ukraine_Processed %>% 
  filter(City %in% "Lviv")

```

> Explain data wrangling process here (shirley do this) 

Data Used | Brief Description 
----------|------------------
UkraineData | (Raw) Ukraine air quality data 
Ukraine_Processed | (Processed) Ukraine air quality data, with no missing values 
Dnipro | (Processed) Dnipro air quality data, showing PM2.5 and PM10 values 
Lviv | (Processed) Lviv air quality data, showing PM2.5 and PM10 values 

\newpage

# Exploratory Analysis  

```{r Plotting PM2.5, fig.cap = "Comparing PM2.5 in Dnipro and Lviv, Ukraine} 

PM25_PLOT <- ggplot(Dnipro, aes(x = date, y = pm25)) + 
              geom_line()+ 
  scale_x_date(limits = as.Date(c("2022-03-01", "2022-03-31")))+ 
  
print(PM25_PLOT)
  


```

```{r Plotting PM10, fig.cap= "Comparing PM10 in Dnipro and Lviv, Ukraine} 

PM10_PLOT <-ggplot()

```



\newpage

# Analysis



## Question 1: <insert specific question here and add additional subsections for additional questions below, if needed>

## Question 2: 




\newpage

# Summary and Conclusions


\newpage

# References
<add references here if relevant, otherwise delete this section> 
