---
output: 
  pdf_document:
    keep_tex: yes
    fig_caption: yes
    number_sections: yes
geometry: margin=2.54cm
title: "Air Quality in Ukraine post Ukraine-Russia Dispute"
subtitle: "Web address for GitHub repository"
author: "Curly Girlies: Shirley Fontanié, Rachel Gordon, and Julia Weinberg"
fontsize: 12pt
mainfont: Times New Roman

---

\newpage
\tableofcontents 
\newpage
\listoftables 
\newpage
\listoffigures 
\newpage

```{r setup, include=FALSE}
# Set your working directory 

getwd() 

# Load your packages  

library(tidyverse)
library(lubridate)
library(ggplot2)

# Set your ggplot theme

ourtheme <- theme_classic(base_size = 14)+ 
  theme(axis.text = element_text(color="black"), 
        legend.position = "top")
theme_set(ourtheme)
  
# Load your datasets
UkraineData <- read.csv("./RawData/Ukraine_AQIData.csv", stringsAsFactors = TRUE)
```


# Rationale and Research Questions

> Research question: How does air quality in various Ukrainian cities differ before and after the Ukraine-Russia dispute?  
*******
> Rationale: On February 24, 2022, Russia attacked Ukraine. The first city attacked was
Lviv and Dnipro. 
> Kyiv was hit February 24th.  



\newpage

# Dataset Information 

> Explain data process here

```{r Data wrangling} 
Dnipro <- UkraineData %>% 
  

```


\newpage

# Exploratory Analysis 



\newpage

# Analysis



## Question 1: <insert specific question here and add additional subsections for additional questions below, if needed>

## Question 2: 




\newpage

# Summary and Conclusions


\newpage

# References
<add references here if relevant, otherwise delete this section> 
